EvoSwarm: A Cloud-Native Architecture for the Reproducible Execution of Multi-population Optimization Algorithms

### 2020-05-18 draft.tex
Added conclusions and reflowed

There are many advantages of container-based architectures over one
based on virtual machines. The main ones are architectural: it makes
the design of an application gravitate towards a set of decoupled
tasks, that use reactive, event based programming, to interact with
the rest of the application; containers hold micro-services that are
much easier to test, integrate and deploy. It also simplifies
the interaction APIs. Of course, that makes them more economical in a
pay-per-use cloud provider, but that is not the main reason why we
have chosen them in this paper; reproducibility of experiments and a
clear and idempotent description of the architecture through {\tt docker-compose} 
% Be consistent with docker compose. Either {\tt docker-compose} or {\sf docker-compose}, but not docker compose, for instance. - JJ - ok
are the main reasons.

@@

To answer the first research question, we need to evaluate the effect of 
the number of populations and the number of
workers on the scalability of the system, we 
propose  an experiment in which we selected $f_4$ (Skew Rastrigin-Bueche separable) from the
BBOB testbed. This function has been used because it is computationally demanding,
and in higher dimensions has been difficult for PSO \cite{el2009black} and GAs
\cite{nicolau2009application} to solve with the required FEs. As we
are only comparing in terms of scalability, the results from a single function
can be better understood, since there are fewer factors involved.
Following the procedure described in Section~\ref{setup}, we executed 
two sets of experiments with five and ten initial populations,
repeating each experiment 30 times using  1, 2, 4, and 8 workers.  We kept the same 
maximum number of FEs, changing the relevant parameters for this. 
See Tables \ref{tab:params:10} and \ref{tab:params:5} 
for the complete list. 
%
\begin{figure}[h!tbp]
  \centering
  \includegraphics[width=\textwidth]{time_diff}
  \caption{Boxplot of the median time required for the execution of 30 
  instances of function $f_4$, in seconds, different number of workers and populations sizes.  }
  \label{fig:spworker:time}
\end{figure}
%
\begin{table}[]
  \caption{The median of the time required to finish an instance of the function $f_4$, 30 runs
  with different number of workers and populations} % time or rate? - JJ
  \label{tab:time}
  \vspace{0.25cm}
    \centering
    \begin{tabular}{ccrr}
    \hline
              &         & 5 Populations    & 10 Populations \\
     Dimension& Workers & Median of rate & Median of rate  \\
    \hline
          2   & 1w      & 2.96   & 3.15   \\
              & 2w      & 1.84   & 1.72   \\
              & 4w      & 1.22   & \textbf{0.96}   \\
              & 8w      & 1.05   & \textbf{0.59}   \\
    \hline
          3   & 1w      & 6.53   & \textbf{7.76}   \\
              & 2w      & 3.64   & 3.83   \\
              & 4w      & 2.33   & 2.26   \\
              & 8w      & 2.17   &  \textbf{1.36}   \\
    \hline
          5   & 1w      & \textbf{12.33}  & 14.19  \\
              & 2w      & 6.63   & \textbf{7.41}   \\
              & 4w      & 4.25   & 4.34   \\
              & 8w      & 3.67   &  \textbf{2.41}   \\
    \hline
          10  & 1w      & \textbf{31.50}  & 38.30  \\
              & 2w      & \textbf{16.07}  & 19.31  \\
              & 4w      &  9.99   & 10.35  \\
              & 8w      & 8.97   &  \textbf{5.92}  \\
    \hline
          20  & 1w      &  116.35 & 124.51 \\
              & 2w      & \textbf{51.18}  & 63.68  \\
              & 4w      & 32.26  & 33.07  \\
              & 8w      & 28.64  & \textbf{18.25}  \\
    \hline
          40  & 1w      & \textbf{369.89} & 448.89 \\
              & 2w      & \textbf{198.43} & 234.50 \\
              & 4w      & 122.39 &  120.62 \\
              & 8w      & 122.31 & \textbf{63.35} \\
    \hline
  \end{tabular}
\end{table}

@@

In this paper we have presented the EvoSwarm framework that implements
an event-based, cloud native architecture suitable for hosting any
kind of multi-population optimization algorithm. The framework is
based in industry-standard development tools such as Docker and Docker
Compose, making easy to develop and deploy multi-algorithm distributed
experiments.

The set up of this kind of algorithms will have a big influence on the
results, since diversity is important for this kind of algorithms and
its creation and destruction play a big part of the algorithm's
performance. One of the main levers to control it in EvoSwarm is via
the number of workers that will be allocated from the beginning, which
is why we have performed an experiment mainly intended as a proof of
concept, but also devoted to show how two parameters in the framework,
number of workers and number of populations used to initially seed the
system, interact with each other. The conclusion of this experiment is
that there is an interesting interplay, with the most benefit being
obtained with a number of populations equal to the number of workers
or slightly inferior or superior, which brings algorithmic benefit
(experiment taking less time because it needs less evaluations) and
also performance benefits (experiment taking less time because
evaluations are being split between different workers.

% this is not a conclusion. The conclusion is an answer to the
% research question. This works, and works precisely for the reasons
% we say it works. - JJ
% Again, revise search questions and explain answers - JJ
As a conclusion, EvoSwarm is a framework designed and tested according
to industry standards which is able to bring many benefits to the
bioinspired algorithm community, by showing the possibility of
creating new, high performance, mixed algorithms, which can also be
easily deployed to the cloud, or locally to systems that simply need
to have Docker and {\tt docker-compose} installed.

It also opens new possibilities. From the design point of view, it
would be interesting to try and use automatic scaling, instead of
setting the number of workers by hand. This would need a certain
amount of redesign, the most important of which would be using
container orchestration systems such as Kubernetes or Docker
Swarm. Instead of setting a fixed number of workers, only the maximum
amount would need to be established, which could be done directly or
by setting an evaluation budget.

>>>>

There are many advantages of container-based architectures over one
based on virtual machines. The main ones are architectural: it makes
the design of an application gravitate towards a set of decoupled
tasks, that use reactive, event based programming, to interact with
the rest of the application; containers hold micro-services that are
much easier to test, integrate and deploy. It also simplifies
the interaction APIs. Of course, that makes them more economical in a
pay-per-use cloud provider, but that is not the main reason why we
have chosen them in this paper; reproducibility of experiments and a
clear and idempotent description of the architecture through {\tt docker-compose} 
are the main reasons.

@@

To answer the first research question, we need to evaluate the effect of the
number of populations and the number of workers on the scalability of the
system, we propose an experiment in which we selected $f_4$ (Skew
Rastrigin-Bueche separable) from the BBOB testbed. This function has been used
because it is computationally demanding, and in higher dimensions has been
difficult for PSO \cite{el2009black} and GAs \cite{nicolau2009application} to
solve with the required FEs. As we are only comparing in terms of scalability,
the results from a single function can be better understood, since there are
fewer factors involved. Following the procedure described in
Section~\ref{setup}, we executed two sets of experiments with five and ten
initial populations, repeating each experiment 30 times using 1, 2, 4, and 8
workers. We kept the same maximum number of FEs, changing the relevant
parameters for this. See Tables \ref{tab:params:10} and \ref{tab:params:5} for
the complete list.

%
\begin{figure}[h!tbp]
  \centering
  \includegraphics[width=\textwidth]{time_diff}
  \caption{Boxplot of the median time required for the execution of 30 
  instances of function $f_4$, in seconds, different number of workers and populations sizes.  }
  \label{fig:spworker:time}
\end{figure}
%
\begin{table}[]
  \caption{The median of the time required to finish an instance of the function $f_4$, 30 runs
  with different number of workers and populations} % time or rate? - JJ
  \label{tab:time}
  \vspace{0.25cm}
    \centering
    \begin{tabular}{ccrr}
    \hline
              &         & 5 Populations    & 10 Populations \\
     Dimension& Workers & Median of rate & Median of rate  \\
    \hline
          2   & 1w      & 2.96   & 3.15   \\
              & 2w      & 1.84   & 1.72   \\
              & 4w      & 1.22   & \textbf{0.96}   \\
              & 8w      & 1.05   & \textbf{0.59}   \\
    \hline
          3   & 1w      & 6.53   & \textbf{7.76}   \\
              & 2w      & 3.64   & 3.83   \\
              & 4w      & 2.33   & 2.26   \\
              & 8w      & 2.17   &  \textbf{1.36}   \\
    \hline
          5   & 1w      & \textbf{12.33}  & 14.19  \\
              & 2w      & 6.63   & \textbf{7.41}   \\
              & 4w      & 4.25   & 4.34   \\
              & 8w      & 3.67   &  \textbf{2.41}   \\
    \hline
          10  & 1w      & \textbf{31.50}  & 38.30  \\
              & 2w      & \textbf{16.07}  & 19.31  \\
              & 4w      &  9.99   & 10.35  \\
              & 8w      & 8.97   &  \textbf{5.92}  \\
    \hline
          20  & 1w      &  116.35 & 124.51 \\
              & 2w      & \textbf{51.18}  & 63.68  \\
              & 4w      & 32.26  & 33.07  \\
              & 8w      & 28.64  & \textbf{18.25}  \\
    \hline
          40  & 1w      & \textbf{369.89} & 448.89 \\
              & 2w      & \textbf{198.43} & 234.50 \\
              & 4w      & 122.39 &  120.62 \\
              & 8w      & 122.31 & \textbf{63.35} \\
    \hline
  \end{tabular}
\end{table}

@@

In this paper, we have presented the EvoSwarm framework that implements an
event-based, cloud-native architecture suitable for hosting multi-population
optimization algorithms. The framework is based in industry-standard development
tools such as Docker and Docker Compose, making it easy to develop and deploy
multi-algorithm distributed experiments. The model is based on the asynchronous
exchange of messages between stateless functions that react to a continuous
stream of data. We compared the message-driven architecture we presented against
four other proposals for multi-population algorithms found in the literature. We
found that it is best suited for asynchronous multi-population algorithms.

We proposed a new methodology for the reproducible deployment and execution of
multiple experiments by specifying the infrastructure as part of an experiment
definition in both local or cloud environments. We presented the workflow of an
experiment's entire life-cycle viewed from the perspective of developers and
researchers. We tested the deployment on both local and Amazon's Elastic
Container Service.

We did an empirical evaluation of the same framework to learn if the proposed
solution can efficiently improve the scalability and performance of
population-based optimization algorithms.

First, we focused on knowing how scaling is related to the number of populations
and workers in the system. This experiment concludes that there is an
interesting interplay, with the most benefit obtained with a number of
populations near-equal to the number of workers, which brings performance gains.
Also, we found that for lower dimensions, there is an advantage in using fewer
populations.

A second experiment centered on evaluating the algorithmic benefits (experiments
taking less time because they need fewer evaluations) of distributed
multi-algorithm algorithms. We compared the homogeneous and an ensemble of
multi-populations, using Genetic Algorithms (GAs) and Particle Swarm
Optimization (PSO) using the COCO benchmark framework for continuous functions.
We found that the set-up of these algorithms has a significant influence on
their results since diversity is essential for their performance. The ensemble
achieved better results and, in most cases, needed fewer evaluations.

% In this paper, we have presented the EvoSwarm framework that implements
% an event-based, cloud-native architecture suitable for hosting 
% multi-population optimization algorithms. The framework is
% based in industry-standard development tools such as Docker and Docker
% Compose, making it easy to develop and deploy multi-algorithm distributed
% experiments.

% The set up of this kind of algorithms will have a big influence on the
% results, since diversity is important for this kind of algorithms and
% its creation and destruction play a big part of the algorithm's
% performance. One of the main levers to control it in EvoSwarm is via
% the number of workers that will be allocated from the beginning, which
% is why we have performed an experiment mainly intended as a proof of
% concept, but also devoted to show how two parameters in the framework,
% number of workers and number of populations used to initially seed the
% system, interact with each other. The conclusion of this experiment is
% that there is an interesting interplay, with the most benefit being
% obtained with a number of populations equal to the number of workers
% or slightly inferior or superior, which brings algorithmic benefit
% (experiment taking less time because it needs less evaluations) and
% also performance benefits (experiment taking less time because
% evaluations are being split between different workers.

% % this is not a conclusion. The conclusion is an answer to the
% % research question. This works, and works precisely for the reasons
% % we say it works. - JJ
% % Again, revise search questions and explain answers - JJ
% As a conclusion, EvoSwarm is a framework designed and tested according
% to industry standards which is able to bring many benefits to the
% bioinspired algorithm community, by showing the possibility of
% creating new, high performance, mixed algorithms, which can also be
% easily deployed to the cloud, or locally to systems that simply need
% to have Docker and {\tt docker-compose} installed.

This work also opens new possibilities. From the design point of view, it
would be interesting to try and use automatic scaling, instead of
setting the number of workers by hand. This would need a certain
amount of redesign, the most important of which would be using
container orchestration systems such as Kubernetes or Docker
Swarm. Instead of setting a fixed number of workers, only the maximum
amount would need to be established, which could be done directly or
by setting an evaluation budget.
